Gradient Descent for Numerical MAX-CSP

### 2022-05-02 proj/paper/paper.tex
new graph
save paper

\documentclass[a4paper]{article}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{dsfont}
\usepackage{paralist}
\usepackage{epstopdf}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{multicol}
\usepackage[hidelinks]{hyperref}
\usepackage{fancyvrb}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{float}
\usepackage{paralist}
\usepackage[svgname]{xcolor}
\usepackage{enumerate}
\usepackage{array}
\usepackage{times}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{comment}
\usepackage{environ}
\usepackage{times}
\usepackage{textcomp}
\usepackage{caption}
\usepackage[square,numbers]{natbib}

@@

\section{Baseline Solutions}

As metioned above we can formulate our problem as a mixed integer linear program 
with binary constraints. In addition, to the mixed integer formulation, we also create an LP
relaxation of the problem as an approximation of the MILP.
The rationale behind this is that the LP relaxation can be solved efficiently and provide a 
deterministic approximation of the MILP. Although the solution could be arbitrarily bad,
we find that the LP typically finds a reasonable solution in much less time 
relative to the MILP.

\section{Problem with Baseline Solutions}

While modeling the MAX-CSP problem as a MILP is simple solution (at least programming-wise), 
it has a few major issues, namely that depending on the input, the solver time could be exponential in 
the number of constraints. Moreover, even commercial solvers can struggle with large problem sizes, 
either taking large amounts of time to produce any feasible solution or having prohibitive memory requirements.
For our motivating example, we would ideally feed in many labeled queries, 
each producing tens of constraints, hence exponential runtime in the number of constraints greatly 
reduces the usefulness of the solution.

@@

To remedy problem of runtime we propose a new solution based on gradient descent. The high level
idea of the solution is simple, begin with a random weight feasible vector $w$ and then perform 
gradient descent until the local minima is found. More specifically we have,

$$
f(w) = \text{HardTanh}(Aw - \gamma)
$$
Where $ \gamma = -1$ and, 
$$
x\in R, \quad \text{HardTanh}(x) = \begin{cases} -1 \text{ if } x \leq -1\\ 1 \text{ if } x \geq 1\\ x \text{ otherwise}\end{cases}
$$

For gradient updates we start with $\eta = .5$ which decays exponentially every 25 gradient updates. We do a total 
of 500 gradient updates and then back track to find the best $w$ (i.e. the one that satisfies as many constraints as 
possible). We then repeat this process starting with random $w$ until we reach a preset time limit.

\section{Experiments}

We ran our experiments on a server with an AMD Ryzen Threadripper 2950X (16c/32t) processor with 
64GB of RAM. Our MAX-CSP instances are generated from real world datasets, with the number of constraints 
ranging from 7769 to 5994778. For the baseline solutions we 
leverage Gurobi, a commercial solver with an academic license. We choose gurobi because it is consistently 
one of the top performing available solvers \cite{benchmark}.

\begin{tabular}{|l|r|r|r|}
\toprule
Dataset &  \# of Constraints &  \# of Columns &  \% Non-Zero \\
\midrule
         $D_0$ &                   7769 &                 14 &         66.24 \\
         $D_1$ &                  19350 &                 18 &         61.46 \\
         $D_2$ &                 111974 &                 18 &         72.83 \\
         $D_3$ &                 196213 &                 30 &         51.62 \\
         $D_4$ &                3711208 &                124 &         25.09 \\
         $D_5$ &                5994778 &                 22 &         68.58 \\
\bottomrule
\end{tabular}

\section{Discussion}
\section{Future Work}

>>>>

\documentclass[a4paper]{article}
\usepackage{geometry}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{dsfont}
\usepackage{paralist}
\usepackage{epstopdf}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{multicol}
\usepackage[hidelinks]{hyperref}
\usepackage{fancyvrb}
\usepackage{float}
\usepackage{paralist}
\usepackage[svgname]{xcolor}
\usepackage{enumerate}
\usepackage{array}
\usepackage{times}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{comment}
\usepackage{environ}
\usepackage{times}
\usepackage{textcomp}
\usepackage{caption}
\usepackage[square,numbers]{natbib}

@@

\section{Baseline Solution}

As motioned above we can formulate our problem as an  MILP with binary
constraints. This MILP can then be feed into any supported off the shelf solver
to produce an exact optimal answer.  This solution is appealing because,
assuming that the solver is correct, it will give an optimal solution, however
it suffers from a few major drawbacks.  First and foremast, depending on the
input, the solver time could be exponential in the number of constraints.
Moreover we found that even commercial solvers can struggle with large problem
sizes, either taking large amounts of time to produce any feasible solution or
having prohibitive memory requirements.  For our motivating use case, we would
ideally feed in many labeled queries, each producing tens of constraints, hence
exponential runtime in the number of constraints greatly reduces the usefulness
of the solution.

@@

To remedy problem of runtime we propose a new solution based on gradient
descent. The high level idea of the solution is simple, begin with a random
weight feasible vector $w$ and then perform gradient descent until the local
minima is found. More specifically we minimize the following function,

$$
f(w) = \text{HardTanh}(Aw - \gamma)^T\overrightarrow{1}
$$

$$
\text{Where }  \gamma = -1 \text{ and for } x\in R,  \text{HardTanh}(x) = \begin{cases} -1 \text{ if } x \leq -1\\ 1 \text{ if } x \geq 1\\ x \text{ otherwise}\end{cases}
$$

\algrenewcommand\textproc{}
\algrenewcommand{\algorithmiccomment}[1]{\hfill// #1}

\begin{algorithm}[H]
\caption{OptimizeGD($f$, $T$)}
\textbf{Input : } The function $f$ to optimize, a time limit $T$
\textbf{Output : } A local minima $w^*$
\begin{algorithmic}[1]
	\State $w^* \gets \overrightarrow{0}$ 
	\While{current time $ < T$}
		\State $w \gets w \sim \mathds{U}_{[.1,1.1]}^n$ \Comment{Initialize $w$ to a uniform random vector}
		\For{$i =  1,...20$}
			\For{$j =  1,...25$}
				\State $w \gets w - \text{ADAM}(\nabla f(w))$
			\EndFor
			\If{ $f(w) < f(w^*)$}
				\State $w^* \gets w$
			\EndIf
		\EndFor
	\EndWhile
	\\\Return $w^*$
\end{algorithmic}
\end{algorithm}

We use the ADAM optimizer \cite{ADAM_paper} because we emprically found it to perform well,
while giving minimal overhead. 
%TODO add pseduo code?
\section{Experiments}

We ran our experiments on a server with an AMD Ryzen Threadripper 2950X (16c/32t) processor with 
64GB of RAM. Our MAX-CSP instances are generated from real world datasets, with the number of constraints 
ranging from 7.7K to 6.0M. For the baseline solutions we 
leverage Gurobi, a commercial solver with an academic license. We choose gurobi because it is consistently 
one of the top performing available solvers \cite{benchmark}. Both solutions where given a timeout of 2700 seconds 
(corresponding to 24 hours of CPU time).
\begin{table}[ht!]
\caption{Dataset statistics}
\begin{tabular}{|l|r|r|r|}
\toprule
Dataset &  \# of Constraints &  \# of Columns &  \% Non-Zero \\
\midrule
         $D_0$ &                   7.7K &                 14 &         66.24 \\
         $D_1$ &                  19.3K &                 18 &         61.46 \\
         $D_2$ &                 111.9K &                 18 &         72.83 \\
         $D_3$ &                 196.2K &                 30 &         51.62 \\
         $D_4$ &                3.7M    &                124 &         25.09 \\
         $D_5$ &                6.0M    &                 22 &         68.58 \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[ht!]
	\caption{Comparison of MILP vs. Gradient Descent (GD)}
	\includegraphics[width=\textwidth]{./time_series.png}
\end{figure}

\section{Discussion}

In this section we will discuss the experimental results. There are
two major points. We first discuss the quality of solutions produced. 
Next, we discuss scalabilty of solutions.

\subsection{Quality of Solutions}

Much to our surprise, the quality of solutions produced by gradient descent
(GD) was very similar if not better than the MILP solution. We attribute this
to the fact that only one dataset was solved to optimality ($D_2$), in fact, we
let $D_0$ run for over 12 hours and MILP still hadn't found an optimal
solution. This was rather suprising because, $D_0$ has less an order of
magnitude fewer constraints than $D_2$. This underlines up a key trade off
between the solutions, namely, quality of solutions vs. runtime predictability.
While the MILP solution will produce optimal solutions given enough time, it is
very hard to predict how long it will take to produce an optimal solution. On
the otherhand, there are no guarantees that can be made about GD in terms of
the quality of solutions but its runtime is very predictable, scaling roughly
linearly with the number of constraints times the number of columns.

\subsection{Scalability}

This then brings us to our second point which is the scalability of the 
solutions. To begin, from $D_4$ and $D_5$, we can clearly see that 
GD has superior scaling properties when compared to MILP in terms of 
runtime. MILP timed out on both datasets before even begining to refine the 
solutions produced. We attribute this again to the vastly different complexity of the 
two solutions. While GD has linear complexity in the number of constraints, 
MILP has best case polynomial complexity in the number of constraints.  For small 
problems this not an issue, but as the problem size increases MILP quickly becomes
impractical for even approximating the optimal solution.

\section{Future Work}

We identify multiple directions for future work, which fall into 
two categories, optimization techniques, and problem applications.

\subsection{Optimization Techniques}

In this paper we have only taken a small look at the possible 
optimization techniques that could be applied. The most obvious 
variation that could tried is to do stochastic gradient descent. This 
variation could have two major potential benefits. First, it can 
decrease both runtime and memory requirements of the solution further by no longer requiring the 
entire dataset be processed on each gradient update. Second, the stochastic 
version may be able to find better solutions as it is much more likely that it 
could escape worse local minima, instead of relying mostly on having a good starting point.

\subsection{Problem Applications}

We restricted our problem setting to numerical MAX-CSPs with linear
constraints, however in principle there is no reason why this technique could
not be applied to arbitrary constraints. In particular, non-convex constraints
would potentially be solved by applying similar techniques, which current
solvers are not able to handle in an capacity. We believe that it is likely
that modifying the optimzation procedure, such as using stochastic gradient
descent, will likely be required to provide good approximations of such
functions.

\section{Conclusion}

In this work we have taken brief look at a particular kind of CSP, namely, 
numerical MAX-CSP's. We motivated this problem with a real world use case for 
tuning full text search engines. We then proposed a solution based on gradient 
descent and gave preliminary evidence to show that it has some key advantages 
over the baseline MILP based solution of the problem, especially for large problem 
sizes where commercial solvers are too slow to practical. We believe that our solution 
shows promise for finding approximate solutions to problems which were not previously tractable
with traditional MILP based solutions. Future work should continue to explore the 
possible variations and possibly apply similar techniques to related problems.
